Variana Reloaded

### 2022-01-31 papers/variana/notes.tex
Include custom optimizer in VS notes

\title{Global optimizer}
\author{Alexis Roche}
%\date{}

@@

\section{Introduction}

@@

\section{Consistency}

@@

\section{Fixed point}

@@

\section{AEP as relaxed KL~minimization}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=1.7]{universe}
\caption{The Universe}
\label{fig:universe}
\end{figure}

>>>>

\def\x{\mathbf{x}}
\def\th{{\boldsymbol{\theta}}}
\def\n{{\boldsymbol{\nu}}}
\def\u{\mathbf{u}}
\def\m{\mathbf{m}}
\def\v{\mathbf{v}}
\def\f{\mathbf{f}}
\def\E{\mathbb{E}}
\def\g{\mathbf{g}}
\def\h{\mathbf{h}}
\def\A{\mathbf{A}}
\def\B{\mathbf{B}}

\title{Variana Reloaded}
\author{Alexis Roche}
%\date{}

@@

\section{Variational sampling recap}

Consider:
\begin{itemize}
\item a reference distribution~$\pi(x)$ (from which we can sample)
\item a factor~$f(x)$ so that $p\equiv \pi f$ is a distribution of interest
\item an approximating factor~$g$ searched under the exponential form:
  $$
  g(x) = e^{\theta^\top \phi(x)},
  $$ where $\phi$ is a known vector-valued function and $\theta$ needs
  to be determined.
\end{itemize}

VS is a stochastic method to approximate the parametric factor~$g$ that minimizes the generalized Kullback-Leibler divergence:
$$
D(\pi f\|\pi g)
= \int \pi (f \log \frac{f}{g} + g - f)
$$

It does so by sampling independent points $(x_1,\ldots,x_n)$ from~$\pi$ and minimizing a Monte Carlo approximation to the KL~divergence, leading to the surrogate criterion:
$$
\tilde{D}(\theta) = \frac{1}{n} \sum_i \ell[ f(x_i), g_\theta(x_i) ],
$$
where $\ell(f, g)$ is the loss function associated with the KL~divergence:
$$
\ell(f, g) \equiv f \log \frac{f}{g} - f + g
$$

We have proved in our initial paper from 2013 that the resulting estimator of the integral, or moment, $I(\phi) \equiv \int \pi f \phi$, is asymptotically unbiased with variance matrix:
$$
\Sigma_n = \frac{1}{n} \int \pi (f - g_\star)^2 \phi \phi^\top,
$$
where $g_\star$ is the genuine (intractable) KL~minimizer:
$$
g_\star \equiv \arg\min_g D(\pi f\|\pi g).
$$

This makes VS fundamentally different from importance sampling, for which the moment is directly estimated via:
$$
\hat{\mu} \equiv \frac{1}{n} \sum_i f(x_i)\phi(x_i),
$$
yielding an unbiased estimator with asymptotic variance:
$$
\Sigma_n^{IS} = \frac{1}{n} \left[
\int \pi f^2 \phi \phi^\top
- I(\phi) I(\phi)^\top
\right]
$$

In particular, VS has the key consistency property unlike IS to be exact when the target factor lies in the approximating family.

\section{Variance upper bound}

Intuitively, the closer the target factor~$f$ to the approximating family, the smaller the VS~asymptotic variance. To show that formally, we would like to upper bound the mean square error $\int\pi(f-g)^2$ by some increasing function of the KL~divergence $D(\pi f\|\pi g)$.

I have been bumping into this issue for years... only to realize now (January 2022) that such an upper bound does exist and is actually pretty simple to exhibit. We will need a straightforward intermediate result:

\paragraph{Alexis theorem:} if $f:\Omega\in\mathbb{R}\to \mathbb{R}$ is a smooth convex function from some open interval reaching its minimum in $a\in\Omega$, assuming for simplicity that $f(a)=0$, and such that the second derivative $f''$ is a decreasing function, then we have that:
$$
\forall x \leq a,
\qquad
f(x) \geq \frac{1}{2} f''(a) (x - a)^2 .
$$

The proof is straightforward from the fact that $f''(x) \geq f''(a)$ since $f''$ is decreasing, then we just need to integrate this relationship twice and take into account that $f'(a)=f(a)=0$.

Next, let us consider the loss function $\ell(f,g)=f\log(f/g)+g-f$ as a function of $g$. We easily see that:
$$
\frac{\partial\ell}{\partial g} = 1 - \frac{f}{g},
\qquad
\frac{\partial^2\ell}{\partial g^2} = \frac{f}{g^2},
$$
implying that the loss is not only convex in~$g$ but also staisfies the conditions of Alexis's theorem with $a=f$. Hence,
$$
\forall g\leq f,
\qquad
\ell(f, g) \geq \frac{1}{2f} (g - f)^2. 
$$

But we can also consider the loss as a function of~$f$ at fixed~$g$. We have:
$$
\frac{\partial\ell}{\partial f} = \log(f),
\qquad
\frac{\partial^2\ell}{\partial g^2} = \frac{1}{f},
$$
which means that Alexis's theorem can be applied again, this time with $a=g$, yielding:
$$
\forall f\leq g,
\qquad
\ell(f, g) \geq \frac{1}{2g} (f - g)^2. 
$$

Concatenating both results, we see that for all $(f, g)$:
$$
(f-g)^2 \leq 2 \max(f, g)\ell(f, g).
$$

Integrating this over~$\pi$, we get this long-waited inequality:
$$
\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g).
$$

This tells us that, as intuition suggests, the mean squared error between factors smoothly goes to zero as the KL~divergence vanishes. But this holds under the mild condition that both $f$ and $g$ remain bounded. This practically rules out the naive application of VS to unbounded importance weight functions, hence the need to perform VS within a constrained framework such as Expectation Propagation.

Note that, if~$f$ is such that all the moments $\int\pi f \phi$ exist, the same goes for $g_\star$ in virtue of the moment matching condition, implying under very mild conditions on the approximating family, that $g_\star$ is upper bounded (easy to see in the Gaussian case, which is BTW the only one we have considered in practice so far.

The above inequality gives a variance upper bound on the zero-order moment (the normalizing constant). More generally, consider a positive real $\phi$. The above result implies straighforwardly:
$$
(f-g)\phi^2
\leq 2 \max(f\phi, g\phi) \ell(f\phi, g\phi)
= 2 \max(f\phi, g\phi) \phi\ell(f, g)
= 2 \max(f\phi^2, g\phi^2) \ell(f, g)
$$
where we use the fact that $\ell(f\phi,g\phi)=\phi\ell(f,g)$. A more general inequality follows by integration, which I would happily call Alexis' inequality:
$$
\int \pi (f-g)^2\phi^2 \leq 2\max(\|f\phi^2\|_\infty, \|g\phi^2\|_\infty) D(\pi f\|\pi g).
$$

This yields the same kind of variance upper bound for any moment estimator we have.

\section{Custom optimizer}

In the following, we consider the canonical problem of minimizing $D(p\|q)$ for some target~$p$, wrt some distribution log-quadratic distribution~$q$. I wrote the following section at Covii in Spring~2019. The goals is to have a scalable optimizer for VS, and the core idea is to re-parameterize the approximating family locally so as to implement~VS using a first-order gradient descent rather than a quasi-Newton method, which requires maninpulations on some Hessian approximation in one way or another and may thus be challenging in very large dimension.

\subsection{Basis change}

Consider minimizing the KL divergence via a gradient descent and consider a single gradient update. If we cannot afford to compute the Hessian (and to invert it), the gradient update relies on an arbitrary metric and may be rather ineffective. An important property of the KL divergence with exponential families is that its Hessian is independent from the target distribution and is associated with a dot product on the linear space spanned by~$\phi$ (which is a subspace of polynomials of order~2 for the factorial Gaussian family):
$$
H = \int q(\x) \phi(\x) \phi(\x)^\top d\x 
\quad \Rightarrow \quad
H_{ij} = \langle \phi_i , \phi_j \rangle .
$$

A natural idea is then to pick an orthonormal basis to use temporarily for a single gradient update as the divergence Hessian is scalar in such a basis.

Can we easily compute such as basis? The answer is yes. Assume for now and for simplicity that $\int q = 1$. In one dimension, it is easy to find an orthonormal basis:
$$
\phi_0(x) = 1,
\quad \phi_1(x) = \frac{(x-m)}{s},
\quad \phi_2(x) = \frac{1}{\sqrt{2}}\left[\frac{(x-m)^2}{s^2} - 1\right].
$$

It is straightforward to see that $\|P_0\|=1$, $\|P_1\|=1$, $\langle P_0, P_1 \rangle=0$ and $\langle P_0, P_2 \rangle=0$. The reason why $\langle P_1, P_2 \rangle=0$ is that it subtracts two centered moments of order~1 and~3, respectively, which both vanish because~$q$ is symmetric. Finally, note that:
$$
\phi_2(x)^2 = \frac{1}{2} 
\left[\frac{(x-m)^4}{s^4} - 2 \frac{(x-m)^2}{s^2} + 1 \right],
$$
hence, given that the 4th-order centered moment of a Gaussian is~3 (which can be seen using the $\chi_1^2$ distribution), we have:
$$
\E_q [\phi_2^2(x)] = \|P_2\|^2 = \frac{1}{2} (3 - 2 + 1) = 1.
$$

In multiple dimensions, we do:
$$
\phi_0(\x) = 1, 
\quad
\phi_{1i}(\x) = \frac{(x_i-m_i)}{s_i},
\quad
\phi_{2i}(\x) = \frac{1}{\sqrt{2}}\left[\frac{(x_i-m_i)^2}{s_i^2} - 1\right].
$$
This defines an orthonormal basis because the components of~$\x$ are mutually independent under the factorial distribution~$q(\x)$, hence for different basis vectors: 
$$
\forall (i,j), \quad i\not= j, \quad
\langle \phi_{ki}, \phi_{\ell j} \rangle = \E_q[\phi_{ki}]\E_q[\phi_{ki}] = \langle \phi_{ki}, 1 \rangle \langle P_{\ell j}, 1 \rangle = 0.
$$
Moreover, when it comes to dot products involving the same component, the one-dimensional results apply.

In the general case where~$q$ does not integrate to one, we simply need to divide each basis vector by $Z=\int q_0$ to fulfill the unit norm conditions. Alternatively, we may leave the basis as is and later remember that the Hessian is $Z\text{Id}$.

\subsection{Iteration}

So we have an orthonormal basis that is very easy to compute! Now, consider decreasing the KL divergence $D(p\|q)$ starting from some current guess~$q_0$. To this end, we may compute an orthonormal basis associated with~$q_0$ and perform one Newton iteration in that basis. More specifically, we search for a new fit under the form:
$$
q_\th(\x) = q_0(\x) e^{\th^\top \phi(\x)},
$$
where~$\phi$ is the orthonormal basis. Clearly, the initial parameter is $\th_0=0$. The Newton update yields:
$$
\th = \gamma \f,
\qquad
\f = -\frac{1}{Z} \u,
\quad
\u \equiv \nabla D(\th_0) = \int (q_0 - p)\phi,
$$
where $\gamma$ is a given step size ($\gamma=1$ for a genuine Newton iteration) and~$Z$ is there for the Hessian correction.

The gradient is generally intractable as it depends on~$p$, but we can approximate it via random sampling or quadrature. There are essentially two options. One is:
$$
\u \approx \int q_0\phi - {\cal I}(p\phi),
$$
where~${\cal I}$ denotes a numerical integration operator. The other option is:
$$
\u \approx {\cal I}[(q_0- p)\phi].
$$
This is where the variational sampling story may start. Option~1 amounts to a direct moment approximation. Option~2, which corresponds to our initial variational sampling idea, is perhaps less intuitive as it ignores the fact that we can compute the ``fitted moment'' exactly, but is arguably better because the error is smaller whenever~$q_0$ is close enough to~$p$. Hence, for a near Gaussian target, we expect less numerical instabilities. Note that, at this level of description, there is no sampling kernel involved -- this is a specific detail of numerical integration.

So, our gradient update takes on the simple form $\th = \gamma\tilde{\f}$, but this is computed in a moving coordinate system since the orthonormal basis changes on each iteration. Is it easy to compute the `natural' parameters of $q_\th=q_0\exp(\th^\top \phi)$ from the moving coordinates~$\th$? Yes, it is.

First, consider the right hand side and try to do the identification:
$$
e^{\th^\top \phi} = \kappa e^{-\frac{(\x-\mu)^2}{2\sigma^2}},
$$
hence, by taking the log:
$$
\log\kappa - \frac{1}{2} \sum_i \frac{(x_i-\mu_i)^2}{\sigma_i^2}
= 
\theta_0 + \sum_i \theta_{1i}\frac{x_i-m_i}{s_i}
+ \frac{1}{\sqrt{2}} \sum_i \theta_{2i}\big[ \frac{(x_i-m_i)^2}{s_i^2} -1 \big]. 
$$
We may differentiate this wrt $x_i$, yielding: 
$$
\frac{x_i-\mu_i}{\sigma_i^2} 
= \frac{\theta_{1i}}{s_i} + \sqrt{2} \theta_{2i} \frac{x_i-m_i}{s_i^2}
= \frac{\sqrt{2} \theta_{2i}}{s_i^2} \big[x_i - m_i + \frac{s_i\theta_{1i}}{\sqrt{2} \theta_{2i}} \big].
$$
By identification, we find:
$$
\sigma_i^2 = - \frac{s_i^2}{\sqrt{2}\theta_{2i}},
$$
and:
$$
\mu_i = m_i - \frac{s_i\theta_{1i}}{\sqrt{2}\theta_{2i}} = m_i + \frac{\sigma_i^2}{s_i}\theta_{1i}.
$$

To find $\kappa$, we note that the integral of $\th^\top \phi$ wrt~$q_0$ is equal to its dot product with~$\phi_0\equiv 1$, which is $\theta_0 \|\phi_0\|^2=\theta_0 Z$ since the basis $\phi/\sqrt{Z}$ is orthonormal. The constant $Z$ cancels out as it appears on both sides, and we find, using the K\"onig formula:
$$
\log \kappa = \theta_0 + \frac{1}{2}\sum_i \frac{s_i^2 + (\mu_i-m_i)^2}{\sigma_i^2}.
$$

We also need to get the parameter~$\th_0$ of~$q_0$, which act as an offset. For this, we just do $\kappa=K_0$, $\mu_i=m_i$, and $\sigma_i=s_i$, yielding:
$$
\theta_{2i}^0=-\frac{1}{\sqrt{2}},
\quad
\theta_{1i}^0 = 0,
\quad
\theta_0^0 = \log K_0 - \frac{d}{2}.
$$

Therefore, the updated fit is a function $\Psi(q_0,\th)$ of the current fit and parameter variation:
$$
v_{i1} =  \frac{v_{i0}}{1-\sqrt{2}\theta_{2i}},
\quad
m_{i1} = m_{i0} + \frac{v_{i1}}{\sqrt{v_{i0}}} \theta_{1i},
\quad
\log K_1 = \log K_0 + \theta_0 + \frac{1}{2} \left[\sum_i \frac{v_{i0} + (m_{i1}-m_{i0})^2}{v_{i1}} - d\right].
$$

\section{Naive VS}

What the variance inequalities essentially prescribe is that the target factor~$f$ be a fast decreasing bounded function (it should not have heavier tails than polynomial in the Gaussian case). In practice, this constraint is easily fulfilled, for instance if~$f$ is a posterior distribution arising from an exponential prior and a bounded likelihood function. Other compliant factors can be derived from this case, as we will see. However, importance weighting functions resulting from distribution ratios are ruled out.

Consider a Bayesian posterior distribution of the form $p=\pi L$, where $\pi$ is a prior and $L$ a likelihood function. A natural approach would be to apply VS with $\pi$ as the sampling distribution and $f=p/\pi=L$ as the factor to be approximated. This means that we are trying to approximate the likelihood function by a Gaussian function in the context of the prior. Makes sense. This is the canonical approach exposed in my previous publications on VS.

However, there are some issues with that. Firstly, the likelihood might not be fast-decreasing (this is the case in image registration), meaning that VS convergence is doubtful. It may not converge at all, or convergence may be super slow... But, even if $L$ is a nice friendly sharply peaked function, sampling from the prior is intuitively not the most efficient thing to do as most simulated points will have low likelihood.

This is what we may call the ``grasp all, loose all'' effect: unless the factor is stricly Gaussian, which never happens in practice (and if it did, VS would not be useful), the VS~variance blows up as the context gets wider. In a narrow context, i.e. $\pi$ is sharp wrt $f$, the factor doesn't vary much within the sampled region, hence it can be very well approximated by a Gaussian (think of the fact that $\log f$ may be close to a quadratic function in this small neighborhood by Taylor's expansion). Such an approximation is bound to be less accurate as the context gets wider.

\section{Variance reduction}

So, how can we make VS more stable?

Instead of trying to solve the problem in one shot, maybe we can do it iteratively. Assume we have a current factor approximation $g_n$. How can we improve it?

Perhaps a natural idea, at least this is among the first that came to my mind when I started to work on VS, is to apply VS sampling from~$\pi g_n$ to the factor $f/g_n$. In my initial framework, it means approximating the target $p=\pi f$ by the distribution $q=\pi g$, using the current approximation $q_n$ as a sampling distribution and $p/q_n$ as the target importance weight function.

But we know that this is a bad idea because $p/q_n=f/g_n$ may not be upper bounded, hence VS is not even guaranteed to converge. In fact, it will diverge if $p$ is a heavy-tailed distribution, which is impossible to know in advance.

A more reasonable thing to do is to note that, since $g_n$ approximates $f$, the combined function $g_n^{1-\alpha}f^\alpha$, for some $0< \alpha\leq 1$, yields an even better approximation. If $g_n$ is far off, it would be a little less far off if we pick a small~$\alpha$, and much closer if we pick a value close to one. If $g_n$ is already a good fit, it shouldn't matter too much which $\alpha$ we pick, at least it should have less impact.

This simple remark suggests that the objective $D(\pi f\|\pi g)$ could be proxied by another KL divergence, $D(\pi g_n^{1-\alpha}f^\alpha \| \pi g)$, yielding the following iteration:
$$
g_{n+1} = \arg\min_g D(\pi g_n^{1-\alpha}f^\alpha \| \pi g)
$$

We will soon realize that this is heavily related to the minimization of the $\alpha$-divergence $D_\alpha(\pi f\|\pi g)$, and also has a strong connection with the Expectation Propagation algorithm, more specifically its ``averaged'' variant (AEP).

But, for the time being, let us see why this is interesting for VS.

\section{Scaling}

We may interpret the KL divergence $D(\pi f\|\pi g)$ as a divergence between the distributions $\pi f$ and $\pi g$, but also as a divergence between the factors $f$ and $g$ in the context defined by~$\pi$. This second interpretation opens a new perspective on~VS, and enables generalizations. For instance, we may consider the more general divergence:
$$
\delta_\varepsilon (f, g) \equiv
D(\pi f^\varepsilon \|\pi g^\varepsilon),
$$
where $\varepsilon$ is a positive scaling factor. The intuition behind this generalization is that the scaled target factor $f^\varepsilon$ gets flatter as $\varepsilon$ approaches zero, hence the VS estimator should have lower variance considering that the approximating family contains constant functions. On the other hand, applying a scaling factor~$\varepsilon$ smaller than one amounts to ``shrinking'' the context with respect to the factor, hence biasing the divergence minimizer (the VS target) towards the particular context. This yields a typical bias/variance tradeoff.

Could the Expectation Propagation (EP) solve this?

$$
p = \pi \prod_{k=1}^n f^\varepsilon,
\quad
\varepsilon = \frac{1}{k}.
$$

Say current approx is $g^\epsilon$ for all factors. An EP update is to fit the factor $f^\varepsilon$ in the cavity $c=\pi g^{1-\varepsilon}$ by minimizing $D(cf^\varepsilon\|cg^\varepsilon)$ over a new $g$. This can be approximated using VS, with $c$ as the context, and $f^\varepsilon$ as the target factor. This is arguably an easier problem than the original one.

\section{Expectation Propagation}

@@

\subsection{Consistency}

@@

\subsection{Fixed point}

@@

\subsection{AEP as relaxed KL~minimization}

@@

%\begin{figure}[h!]
%\centering
%\includegraphics[scale=1.7]{universe}
%\caption{The Universe}
%\label{fig:universe}
%\end{figure}
